Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-19 article.Rmd
Minor changes to first example

The possibilities for inferring gender from names depends on two things. First, it needs a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by governments, though their compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year or range of years, since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to formulate more accurate predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Ben Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as provided a significant correction to predictions based on Social Security Administration data.

@@

These two datasets (and any future datasets to be added to the package) thus contain a simple table of data. ^[The format of this data departs from the more generally desirable "tidy data" framework proposed by Hadley Wickham for performance reasons. By keeping the data in a "wider" format instead of a tidy format as defined by Wickham, the function is able to make its predictions faster [Wickham 2014b].] The columns contain the `name`, `year`, and number of `female` and `male` uses of that name in a particular year.

@@

The method for predicting gender from a name using the package's datasets is simple. Let's begin by assuming that we want to predict the gender of someone named Sidney who was born in 1935 using the Social Security Administration dataset. Because the dataset contains a list of names for each year, we can simply lookup the row for Sidney in 1935. Using the [dplyr](https://github.com/hadley/dplyr) package for R, this can be expressed in a simple grammar of filtering by name and year.

@@

Next we can sum up the male and female columns ("summarize" in dplyr's vocabulary) and calculate the proportions of female and male uses of "Sideny" during that decade.

@@

In other words, for the decade of the 1930s, we can calculate that there is a `r gender("sidney", year = c(1930, 1939))$proportion_male %>% percent()` percent chance that person named Sidney was male. This is roughly the same as the probability we calculated above for just 1935, but our method also returns the figures it used to calculate those probabilities: 1,067 total instances of "Sidney" in 1935 versus 10,110 instances for the decade as a whole.

>>>>

The possibilities for inferring gender from names depends on two things. First, it needs a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by governments, though their compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year or range of years, since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to formulate more accurate predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Benjamin Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as provided a significant correction to predictions based on Social Security Administration data.

@@

These two datasets (and any future datasets to be added to the package) thus contain a simple table of data.^[The format of this data departs from the more generally desirable "tidy data" framework proposed by Hadley Wickham for performance reasons. By keeping the data in a "wider" format instead of a tidy format as defined by Wickham, the function is able to make its predictions faster [Wickham 2014b].] The columns contain the `name`, `year`, and number of `female` and `male` uses of that name in a particular year.

@@

The method for predicting gender from a name using the package's datasets is simple. Let's begin by assuming that we want to predict the gender of someone named Sidney who was born in 1935 using the Social Security Administration dataset. Because the dataset contains a list of names for each year, we can simply lookup the row for Sidney in 1935. Using the [dplyr](https://github.com/hadley/dplyr) package for R, which provided a grammar for data manipulation, this can be expressed with the action "filter":

@@

Next we can sum up the male and female columns ("summarize" in dplyr's vocabulary) and calculate the proportions of female and male uses of "Sidney" during that decade.

@@

In other words, for the decade of the 1930s, we can calculate that there is a `r gender("sidney", year = c(1930, 1939))$proportion_male %>% percent()` percent chance that person named Sidney was male. This is roughly the same as the probability we calculated above for just 1935, but our method also returns the figures it used to calculate those probabilities: 1,067 instances of "Sidney" in 1935 versus 10,110 total instances for the decade as a whole.
